Active Learning for SAT Solver Benchmarking

### 2022-10-08 paper/paper.tex
section3.3

@@

\subsection{Predicting Rankings}
\label{sec:main1}
In order to predict a ranking of solver $\hat{a}$ amidst its competitors, we need to infer its performance on all instances, even the ones on which we have not evaluated it.
To do so, there are generally two notions of this problem, i.e., regression and classification.
A regression model tries to predict the \mbox{PAR-2} runtime of $\hat{a}$ on all instances based on the samples that have been selected.
Then, the ranking is induced by the average predictions.
Alternatively, a classification model can be used to predict some discrete runtime labels of $\hat{a}$ on all instances.
Similarly, the average runtime label determines the ranking.

Preliminary experiments, as well as research in portfolio solvers~\cite{NgokoCT19,CollauttiMMO13}, have shown however that directly predicting runtimes without any transformation does not work well in practice.
%JB: Collauti at el. still predict actual (though normalized) runtimes rather than discrete labels, they only heavily discretize runtimes when using them as features
%TF: added "without any transformation"
%In our experiments, the mean squared error was within the same magnitude as the values to be predicted.
In contrast to that, discretizing runtimes and using those labels in the model's training and prediction yields good results (cf.~Section~\ref{sec:eval}).
The following paragraphs motivate why it makes sense to discretize runtimes.
%JB: we should probably cut this a bit
%TF: I've commented out a few sentences but we probably need to cut more.

We use a function $\gamma\!: {\mathcal{A} \times \mathcal{I}} \rightarrow \left\lbrace 1, \dots, k \right\rbrace$ to transform runtime observations.
In our experiments, this function is a clustering algorithm.
Note that this transformation is done per instance to adapt to different instance difficulties. 
Also, we define ${\gamma\!\left(a, i\right) := k}$ iff $r\!\left(a, i\right) = \,\perp$ for all $a \in \mathcal{A}$ and $i \in \mathcal{I}$.
To determine a ranking based on those labels, we define
\begin{equation}
  \operatorname{score}\!\left(a\right) := \frac{1}{|\mathcal{I}|} \sum_{i \in \mathcal{I}} w_{a,i} \cdot \left(\gamma\!\left(a, i\right) - 1\right) \enspace \textrm{,}
  %JB: does the "-1" provide any benefits, or does it only shift scores?
  %TF: it is used to resemble the 0-indexed implementation. I'm also not happy with this, but using \lbrace 0, \dots, k - 1 \rbrace does also look ugly.
  \label{eq:rankingeq}
\end{equation}
with $w_{a,i} := 2$ (PAR-\emph{2} penalty factor) if $\gamma\!\left(a, i\right) = k$ else $w_{a,i} := 1$.
To be applicable to $\hat{a}$, we extend \emph{score} by replacing $\gamma\!\left(a, i\right)$ by the model's predictions $f\!: \left(\mathcal{A} \cup \left\lbrace \hat{a} \right\rbrace\right) \times \mathcal{I} \rightarrow \left\lbrace 1, \dots, k \right\rbrace$.
%JB: as "$\operatorname{score}$" is not visibly different to the rest of the text, we should use some other way of highlighting
%TF: replaced it with emph{score} for now

Turning a regression into a classification problem, discretization is a useful tool.
However, it does not provide much value if the resulting cluster labels are not predictive of the PAR-2 ranking we seek to predict.
In our work, we use a hierarchical-clustering approach with a single-link criterion for $k = 3$.
For the SAT Competition~2022 Anniversary Track instances~\cite{sat2022}, this ranking approach is capable of correctly deciding for almost all pairs of solvers which one is faster (about \SI{97.45}{\%}; $\sigma = \SI{3.68}{\%}$).
Moreover, it decides all solvers correctly if the difference in the cluster-label-ranking scores between two solvers is at least \SI{0.11}{}.
%Refer to Table~\ref{tab:discreteranking} for a comparison with the PAR-2 ranking.
The Spearman correlation~\cite{de2016comparing} of cluster-label and PAR-2 ranking is about \SI{0.988} (in comparison to a perfect value of \SI{1}).
This shows that monotonicity and score distances between ranks align quite well between both ranking methods.

@@

\subsection{Instance Sampling}
\label{sec:main2}
Our instance-sampling strategies (also refer to Alg.~\ref{algALBenchmark}; line 4) can be divided into two categories.
First, we present possible baseline approaches including existing work on instance-based selection strategies~\cite{Bossek021a,GentHJKMNN14,MatriconAFSH21}.
%JB: approaches under "Second" are also at least "inspired by existing work" and probably also pick "interesting" instances, so distinction between the two categories of approaches not entirely clear
%TF: made it clearer. First = instance-based methods, second = model-based methods
Second, we show promising machine-learning-based sampling strategies that make use of the model's label-prediction function $f$ and are inspired by existing work within the realms of active learning~\cite{settles2009active}.

@@

\subsection{Stopping Criteria}
\label{sec:main3}
Similar to the instance-selection strategies, we can also divide stopping criteria into baseline and active-learning-based approaches.
Also, refer to~Alg.~\ref{algALBenchmark}, line~3.
While a fixed subset-size criterion, e.g., only sample \SI{20}{\%} of all available instances, is an example of the former, a stopping criterion based on a Wilcoxon signed-rank test or the convergence of the model's predictions is an example for the latter.

The Wilcoxon stopping criterion stops the iterative active-learning algorithm when we are on average certain enough that the predicted runtime labels are sufficiently different in comparison to the other solvers.
To be more specific, we stop when
\begin{equation}
  W_{\hat{a}} := \frac{1}{\left\lvert \mathcal{A} \right\rvert} \sum_{a \in \mathcal{A}} \operatorname{wilcoxon}_{\alpha}\!\left(\gamma\!\left(a, \mathcal{I}\right), f\!\left(\hat{a}, \mathcal{I}\right) \right)
\end{equation}
drops below a certain threshold.
%JB: if this is info gain regarding runtimes, doesn't Eq. (4) mean that we would prefer instances with low entropy of runtime distribution after considering the new solver, i.e., instances that are not very discriminatory?
%TF: I don't think so. We want to maximize the reduction in entropy of the runtime labels. Low entropy afterward -> we know that the solver behaves similarly to existing ones.
Thereby, \emph{wilcoxon} denotes the $p$-value of a Wilcoxon signed-rank test of the runtime-label distributions and $\gamma\!\left(a, \mathcal{I}\right)$ and $f\!\left(\hat{a}, \mathcal{I}\right)$ are the natural set extensions on $\gamma$ and $f$ respectively.
%JB: "set extension" might be rather technical and hard to understand; also, why "expected"? in the sense of "straightforward"?
%TF: I already saw this expression in other papers. I'm not sure how to say the same thing clearer in the same amount of space.
To improve the stability of this criterion, we use an exponential moving average
\begin{equation}
  W_{\exp}^{\left(m\right)} := \beta W_{\hat{a}} + \left(1 - \beta\right) W_{\exp}^{\left(m - 1\right)} \enspace \textrm{,}
\end{equation}
whereby $m$ denotes the count of iterations within Alg.~\ref{algALBenchmark} and $W_{\exp}^{\left(0\right)} := 1$.

@@

\subsection{Optimization Goal}
\label{sec:goal}
As already stated in the introductory section, this work addresses the \emph{New-Solver Problem} (cf.~Def.~\ref{def:new-solver-problem}).
By running Alg.~\ref{algALBenchmark}, we obtain a prediction model $\mathcal{M}$ that provides us with a scoring function \emph{score}.
Also, it outputs the fraction of runtime that is needed to arrive at its conclusion.
First and foremost, our goal is to provide the engineer of new SAT solvers with an accurate ranking.
We define the \emph{ranking accuracy} $O_{\operatorname{acc}} \in \left[0, 1\right]$ (higher is better) by the fraction of pairs $\left(\hat{a}, a\right)$ for all $a \in \mathcal{A}$ that are decided correctly by the given ranking (cf. Alg.~\ref{alg:eval}).
For now, we exclude the equality of solvers since data shows that the PAR-2 scores of all our solvers are different.
So, possible ranking decisions may be, for example, solver $a$ is better than $b$ or $b$ is better than $a$.
The ranking accuracy is affected by whether this decision is correctly made.
Matricon et. al.~\cite{MatriconAFSH21} also evaluate their approach by the fraction of solver pairs that is correctly decided.
Second, we also have to optimize for runtime.
The \emph{fraction of runtime} that the algorithm needs to arrive at its conclusion is denoted by $O_{\operatorname{rt}} \in \left[0, 1\right]$ (lower is better).
This metric puts the runtime summed over the sampled instances in relation to the runtime summed over all instances in the dataset (cf. Alg.~\ref{alg:eval}).
Overall, we want to find an approach that maximizes
\begin{equation}
  O_\delta := \delta O_{\operatorname{acc}} + \left(1 - \delta\right) \left(1 - O_{\operatorname{rt}}\right) \enspace \textrm{,}
  \label{eq:opt}
\end{equation} 
whereby $\delta \in \left[0, 1\right]$ allows for linear weighting between the two optimization goals $O_{\operatorname{acc}}$ and $O_{\operatorname{rt}}$.
Plotting the approaches that maximize $O_\delta$ for all $\delta \in \left[0, 1\right]$ on a $O_{\operatorname{rt}}$-$O_{\operatorname{acc}}$-diagram provides us with a Pareto front of the best approaches for different optimization-goal weightings.

\subsection{Evaluation Framework}
\label{sec:evalframe}
To evaluate a concrete instantiation of Alg.~\ref{algALBenchmark} (a concrete choice for all the sub-routines), we perform cross-validation on our set of solvers.
That means that each solver once plays the role of the new solver.
Alg.~\ref{alg:eval} shows this.
Note that the \emph{new} solver in each iteration is excluded from the set of solvers $\mathcal{A}$ to avoid data leakage.
Also, the runtime function $r$ is restricted to $\mathcal{A} \times \mathcal{I}$.

>>>>

Let benchmark instances $\mathcal{I}$, SAT solvers $\mathcal{A}$, and runtimes $r\!: \mathcal{A} \times \mathcal{I} \rightarrow \left[0, \tau\right]$ be given.

@@

\subsection{Runtime Preprocessing and Rank Prediction}
\label{sec:main1}

%In order to predict a ranking of solver $\hat{a}$ amidst its competitors, we need to infer its performance on all instances, even the ones on which we have not evaluated it.
%To do so, there are generally two notions of this problem, i.e, regression and classification.
%A regression model tries to predict the \mbox{PAR-2} runtime of $\hat{a}$ on all instances based on the samples that have been selected.
%Then, the ranking is induced by the average predictions.
%Alternatively, a classification model can be used to predict some discrete runtime labels of $\hat{a}$ on all instances.
%Similarly, the average runtime label determines the ranking.
% MI: skipping this elaboration, since we do not use regression

Preliminary experiments, as well as research in portfolio solvers, have shown however that directly predicting runtimes without any transformation does not work well in practice~\cite{NgokoCT19,CollauttiMMO13}.
%JB: Collauti at el. still predict actual (though normalized) runtimes rather than discrete labels, they only heavily discretize runtimes when using them as features
%TF: added "without any transformation"
In our experiments, the mean squared error was within the same magnitude as the values to be predicted.
In contrast to that, discretizing runtimes and using those labels in the model's training and prediction yields good results (cf.~Section~\ref{sec:eval}).
%The following paragraphs motivate why it makes sense to discretize runtimes.%MI: no further motivation found, commented out
%JB: we should probably cut this a bit
%TF: I've commented out a few sentences but we probably need to cut more.

We use a function $\gamma\!: {\mathcal{A} \times \mathcal{I}} \rightarrow \left\lbrace 1, \dots, k \right\rbrace$ to transform runtime observations.
In our experiments, this function is a clustering algorithm.
Note that this transformation is done per instance to adapt to different instance difficulties. 
Also, we define ${\gamma\!\left(a, i\right) := k}$ iff $r\!\left(a, i\right) = \,\perp$ for all $a \in \mathcal{A}$ and $i \in \mathcal{I}$.
To determine a ranking based on those labels, we define
\begin{equation}
  \operatorname{score}\!\left(a\right) := \frac{1}{|\mathcal{I}|} \sum_{i \in \mathcal{I}} w_{a,i} \cdot \left(\gamma\!\left(a, i\right) - 1\right) \enspace \textrm{,}
  %JB: does the "-1" provide any benefits, or does it only shift scores?
  %TF: it is used to resemble the 0-indexed implementation. I'm also not happy with this, but using \lbrace 0, \dots, k - 1 \rbrace does also look ugly.
  %MI: Why is it so important to have this zero-based?
  \label{eq:rankingeq}
\end{equation}
with $w_{a,i} := 2$ if $\gamma\!\left(a, i\right) = k$ else $w_{a,i} := 1$.
To be applicable to $\hat{a}$, we extend \emph{score} by replacing $\gamma\!\left(a, i\right)$ by the model's predictions $f\!: \left(\mathcal{A} \cup \left\lbrace \hat{a} \right\rbrace\right) \times \mathcal{I} \rightarrow \left\lbrace 1, \dots, k \right\rbrace$.
%JB: as "$\operatorname{score}$" is not visibly different to the rest of the text, we should use some other way of highlighting
%TF: replaced it with emph{score} for now

Discretization would not provide much value if the resulting cluster labels were not predictive of the PAR-2 ranking we seek to predict.
In our work, we use a hierarchical-clustering approach with a single-link criterion for $k = 3$.
For the SAT Competition~2022 Anniversary Track instances~\cite{sat2022}, this ranking approach is capable of correctly deciding for almost all pairs of solvers which one is faster (about \SI{97.45}{\%}; $\sigma = \SI{3.68}{\%}$).
Moreover, it decides all solvers correctly if the difference in the cluster-label-ranking scores between two solvers is at least \SI{0.11}{}.
%Refer to Table~\ref{tab:discreteranking} for a comparison with the PAR-2 ranking.
The Spearman correlation~\cite{de2016comparing} of cluster-label and PAR-2 ranking is about \SI{0.988} (in comparison to a perfect value of \SI{1}).
This shows that monotonicity and score distances between ranks align quite well between both ranking methods.

@@

\subsection{Instance Sampling}
\label{sec:main2}
Our instance-sampling strategies (also refer to Algorithm~\ref{algALBenchmark}; line 4) can be divided into two categories.
First, we present possible baseline approaches including existing work on instance-based selection strategies~\cite{Bossek021a,GentHJKMNN14,MatriconAFSH21}.
%JB: approaches under "Second" are also at least "inspired by existing work" and probably also pick "interesting" instances, so distinction between the two categories of approaches not entirely clear
%TF: made it clearer. First = instance-based methods, second = model-based methods
Second, we show promising machine-learning-based sampling strategies that make use of the model's label-prediction function $f$ and are inspired by existing work within the realms of active learning~\cite{settles2009active}.

@@

\subsection{Stopping Criteria}
\label{sec:main3}
Similar to the instance-selection strategies, we can also divide stopping criteria into baseline and active-learning-based approaches.
Also, refer to~Algorithm~\ref{algALBenchmark}, line~3.
While a fixed subset-size criterion, e.g., only sample \SI{20}{\%} of all available instances, is an example of the former, a stopping criterion based on a Wilcoxon signed-rank test or the convergence of the model's predictions is an example for the latter.

The Wilcoxon stopping criterion stops the iterative active-learning algorithm when we are on average certain enough that the predicted runtime labels are sufficiently different in comparison to the other solvers.
To be more specific, we stop when
\begin{equation}
  W_{\hat{a}} := \frac{1}{\left\lvert \mathcal{A} \right\rvert} \sum_{a \in \mathcal{A}} \operatorname{wilcoxon}_{\alpha}\!\left(\gamma\!\left(a, \mathcal{I}\right), f\!\left(\hat{a}, \mathcal{I}\right) \right)
\end{equation}
drops below a certain threshold.
%JB: if this is info gain regarding runtimes, doesn't Eq. (4) mean that we would prefer instances with low entropy of runtime distribution after considering the new solver, i.e., instances that are not very discriminatory?
%TF: I don't think so. We want to maximize the reduction in entropy of the runtime labels. Low entropy afterward -> we know that the solver behaves similarly to existing ones.
Thereby, \emph{wilcoxon} denotes the $p$-value of a Wilcoxon signed-rank test of the runtime-label distributions and $\gamma\!\left(a, \mathcal{I}\right)$ and $f\!\left(\hat{a}, \mathcal{I}\right)$ are the natural set extensions on $\gamma$ and $f$ respectively.
%JB: "set extension" might be rather technical and hard to understand; also, why "expected"? in the sense of "straightforward"?
%TF: I already saw this expression in other papers. I'm not sure how to say the same thing clearer in the same amount of space.
To improve the stability of this criterion, we use an exponential moving average
\begin{equation}
  W_{\exp}^{\left(m\right)} := \beta W_{\hat{a}} + \left(1 - \beta\right) W_{\exp}^{\left(m - 1\right)} \enspace \textrm{,}
\end{equation}
whereby $m$ denotes the count of iterations within Algorithm~\ref{algALBenchmark} and $W_{\exp}^{\left(0\right)} := 1$.

@@

\subsection{Optimization Goal}
\label{sec:goal}
As already stated in the introductory section, this work addresses the \emph{New-Solver Problem} (cf.~Definition~\ref{def:new-solver-problem}).
By running Algorithm~\ref{algALBenchmark}, we obtain a prediction model $\mathcal{M}$ that provides us with a scoring function \emph{score}.
Also, it outputs the fraction of runtime that is needed to arrive at its conclusion.
First and foremost, our goal is to provide the engineer of new SAT solvers with an accurate ranking.
We define the \emph{ranking accuracy} $O_{\operatorname{acc}} \in \left[0, 1\right]$ (higher is better) by the fraction of pairs $\left(\hat{a}, a\right)$ for all $a \in \mathcal{A}$ that are decided correctly by the given ranking (cf. Alg.~\ref{alg:eval}).
For now, we exclude the equality of solvers since data shows that the PAR-2 scores of all our solvers are different.
So, possible ranking decisions may be, for example, solver $a$ is better than $b$ or $b$ is better than $a$.
The ranking accuracy is affected by whether this decision is correctly made.
Matricon et. al.~\cite{MatriconAFSH21} also evaluate their approach by the fraction of solver pairs that is correctly decided.
Second, we also have to optimize for runtime.
The \emph{fraction of runtime} that the algorithm needs to arrive at its conclusion is denoted by $O_{\operatorname{rt}} \in \left[0, 1\right]$ (lower is better).
This metric puts the runtime summed over the sampled instances in relation to the runtime summed over all instances in the dataset (cf. Alg.~\ref{alg:eval}).
Overall, we want to find an approach that maximizes
\begin{equation}
  O_\delta := \delta O_{\operatorname{acc}} + \left(1 - \delta\right) \left(1 - O_{\operatorname{rt}}\right) \enspace \textrm{,}
  \label{eq:opt}
\end{equation} 
whereby $\delta \in \left[0, 1\right]$ allows for linear weighting between the two optimization goals $O_{\operatorname{acc}}$ and $O_{\operatorname{rt}}$.
Plotting the approaches that maximize $O_\delta$ for all $\delta \in \left[0, 1\right]$ on a $O_{\operatorname{rt}}$-$O_{\operatorname{acc}}$-diagram provides us with a Pareto front of the best approaches for different optimization-goal weightings.

\subsection{Evaluation Framework}
\label{sec:evalframe}
To evaluate a concrete instantiation of Algorithm~\ref{algALBenchmark} (a concrete choice for all the sub-routines), we perform cross-validation on our set of solvers.
That means that each solver once plays the role of the new solver.
Alg.~\ref{alg:eval} shows this.
Note that the \emph{new} solver in each iteration is excluded from the set of solvers $\mathcal{A}$ to avoid data leakage.
Also, the runtime function $r$ is restricted to $\mathcal{A} \times \mathcal{I}$.
